---
title: "Cancer in Scotland"
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
---
# Cancer in Scotland

```{r}
library(readr)
```


```{r}
source(here::here("R/phs_data_info.R"))
source(here::here("R/screening.R"))
# Retrieve helper functions for manipulating bowel cancer screening data

screening_bowel_cancer_takeup_all <- read_csv(here::here(
  "data_clean", phs_screening_bowel_uptake_filepath
))
```
```{r}
theme_set(theme_minimal())
```
# Overview

## Data Quality
https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/deaths-background-information/quality-of-nrs-data-on-deaths

# Analysis

## Death

When a death is registered, it is common for a number of diseases or conditions to
be recorded on the death certificate. National Records of Scotland apply the WHO
rules to determine which of these conditions is the underlying cause of death." - 
[Death Certificates and Coding the Causes of Death](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/deaths-background-information/death-certificates-and-coding-the-causes-of-death)

## Screening

### Scottish Bowel Screening Programme

Commenced in 2007 with a full rollout across all health boards by 2009.

People aged 50-74 years are invited every two years to complete a home bowel screening test. 75 and over can self-refer,

#### Uptake
Based on [Scottish bowel screening programme statistics For the period of invitations from May 2020 to April 2022](https://publichealthscotland.scot/publications/scottish-bowel-screening-programme-statistics/scottish-bowel-screening-programme-statistics-for-the-period-of-invitations-from-may-2020-to-april-2022/)

"The proportion of invites which are returned as a correctly
completed kit (i.e. a positive or negative test result). It is a key measure of the
acceptability of the test to the public."

Healtth Improvement Scotland [Bowel screening standards: August 2023]() 

```{r}
screening_bowel_cancer_takeup_scotland <-get_screening_bowel_cancer_takeup_scotland(screening_bowel_cancer_takeup_all)

#screening_bowel_cancer_takeup_scotland
#screening_bowel_cancer_takeup_all

plot_screening_bowel_cancer_takeup_scotland(screening_bowel_cancer_takeup_scotland)

#plot_bowel
#plot_screening_bowel_cancer_takeup_scotland(screening_bowel_cancer_takeup_all)
```



Bowel Cancer Screening around 2/3 of the people invited via the screening 
programme are taking up the offer.

Females are more likely to return the tests than males.

Conclusion: 

Generally good take up but still room for improvement.

* Continue to enhance communication with the general public making more efforts 
to target males.


# Conclusion

Bowel Cancer Screening is used 