---
title: "Analysis: Waiting Times for Cancer Treatment"
subtitle: "Cancer in Scotland"
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 5
    toc_float: true
editor_options: 
  chunk_output_type: inline
---
```{r setup_waiting_times, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  error = FALSE,
  message = FALSE
)
```
# Waiting Times for Cancer Treatment

## Background

For progressive diseases like cancer early treatment is important to improve the
chances of a better outcome for the patient.

In the publication [Better Cancer Care, An Action Plan](https://www.gov.scot/publications/better-cancer-care-action-plan/), The 
Scottish Government has set targets that once a cancer has been diagnosed 
the patients should start treament within a partcular timeframe.

"...Current standards for cancer waiting times are that 95% of all eligible
patients should wait no longer than 31 or 62 days..."

For this we will be using Public Health Scotland dataset for Cancer Waiting Times.
"Cancer Waiting Times statistics for the 62 day standard for patients urgently 
referred with a suspicion of cancer to first cancer treatment and for the 31 
day standard for patients regardless of the route of referral from date decision 
to treat to first cancer treatment." - Public Health Scotland

Source:  [Cancer Waiting Times](https://www.opendata.nhs.scot/dataset/cancer-waiting-times) [Public Health Scotland]

## Analysis

```{r waiting_times}
library(dplyr)
library(forcats)
library(lubridate)
library(readr)
library(stringr)
library(tidyr)
library(zoo)
```
```{r}
source(here::here("R/cancer_in_scotland.R"))
source(here::here("R/waiting_times.R"))
source(here::here("R/theming.R"))
```

```{r}
# Set custom theme
theme_set(theme_cancer_in_scotland())
```

```{r}
# Read in clean Waiting Times data. N.B. This has the 31 day and 62 day data
# combined
waiting_times_clean <- read_csv("../data_clean/phs/waiting_times/phs_waiting_time_day_standards.csv")
```

```{r}
# skimr::skim(waiting_times_clean)
# View(waiting_times_clean)

waiting_times <- waiting_times_clean %>%
  # add helper variables for visualisation
  mutate(
    quarter_time = zoo::as.yearqtr(quarter_time),
    quarter_new = quarter(quarter_time),
    year_new = as_factor(year(quarter_time)),
    .after = quarter_time
  )
```

```{r}
waiting_times_info <- waiting_times %>%
  distinct(quarter_time) %>%
  summarise(
    count_quarters = n(),
    oldest_quarter = min(quarter_time),
    newest_quarter = max(quarter_time),
    res = newest_quarter - oldest_quarter
  )

# 11 complete years of data
```

We have  **`r waiting_times_info$count_quarters`**  quarters of Waiting Times data 
available ranging from **`r waiting_times_info$oldest_quarter`** to **`r waiting_times_info$newest_quarter`**.
We will combine these to look at complete year totals.

```{r}
# [1] "quarter"                      "quarter_time"
#  [3] "quarter_new"               "year_new"
#  [5] "hb"                           "hbt"
#  [7] "hbtqf"                        "cancer_type"
#  [9] "cancer_type_qf"               "hbqf"
# [11] "referrals_standard"           "referrals"
# [13] "referrals_standard_qf"        "referrals_standard_qf_values"
# [15] "referrals_treated_standard"   "referrals_treated"
# [17] "referrals_treated_qf"         "referrals_treated_qf_values"
```

```{r}
waiting_times_all_cancer_types <- "All Cancer Types"

totals_quarter_31 <- waiting_times %>%
  filter(cancer_type == waiting_times_all_cancer_types) %>%
  filter(referrals_standard == "31day_standard") %>%
  group_by(year_new) %>% # , referrals, referrals_treated) %>%
  summarise(
    total_31_day_referrals = sum(referrals),
    total_31_day_referrals_treated = sum(referrals_treated),
    pct_31_day_treated = (total_31_day_referrals_treated / total_31_day_referrals) * 100
  ) %>%
  ungroup()
```

```{r}
totals_quarter_62 <- waiting_times %>%
  filter(cancer_type == waiting_times_all_cancer_types) %>%
  filter(referrals_standard == "62day_standard") %>%
  group_by(year_new) %>% # , referrals, referrals_treated) %>%
  summarise(
    total_62_day_referrals = sum(referrals),
    total_62_day_referrals_treated = sum(referrals_treated),
    pct_62_day_treated = (total_62_day_referrals_treated / total_62_day_referrals) * 100
  ) %>%
  ungroup()

# Join these two sets of results on year
all_standards_totals <- totals_quarter_31 %>%
  left_join(totals_quarter_62) %>%
  filter(year_new != 2023) # remove partial year

rm(totals_quarter_31, totals_quarter_62)
```

```{r}
all_standards_totals_filtered <- all_standards_totals %>%
  pivot_longer(
    cols = c(
      "total_31_day_referrals",
      "total_31_day_referrals_treated",
      "total_62_day_referrals",
      "total_62_day_referrals_treated"
    ),
    names_to = "referrals_names",
    values_to = "referrals_values"
  ) %>%
  pivot_longer(
    cols = c(
      "pct_31_day_treated",
      "pct_62_day_treated"
    ),
    names_to = "percent_names",
    values_to = "percent_values"
  ) %>%
  # all_standards_totals_filtered <- all_standards_totals_combined %>%
  filter(
    ((referrals_names == "total_31_day_referrals") & (percent_names == "pct_31_day_treated")) |
      ((referrals_names == "total_31_day_referrals_treated") & (percent_names == "pct_31_day_treated")) |
      ((referrals_names == "total_62_day_referrals") & (percent_names == "pct_62_day_treated")) |
      ((referrals_names == "total_62_day_referrals_treated") & (percent_names == "pct_62_day_treated"))
  )
```

### Number of Referrals

```{r}
plot_all_standards <- plot_all_standards_totals_filtered(all_standards_totals_filtered)
do_cis_plot(plot_all_standards, cancer_waiting_times_number_referrals_plot_filepath)
```

The trend is for increasing numbers of referrals being made.
Of concern is between 2021 and 2022 the gap between those referred and 
those treated within the target windows of time is at its largest. 
It is likely that the referral system is still processing a lot of people who 
developed symptoms during the year COVID-19 started (2020) but did not seek or 
were not able to be referred. Is it possible that people are presenting with 
conditions at a more advanced stage of the disease and now need more complex 
diagnostic procedures taking longer to reach the right treatment?

**Suggestion:**

It seems that more resources need to be directed towards improving getting 
treatment underway to undo the immediate delays from COVID-19. If there are any 
administrative processes that could be hastened or improvements in 
information/communication that would be helpful for reducing the wait for patients.

### Are Cancer Waiting Time targets being met?

```{r}
waitiing_times_years_percent_treated <- all_standards_totals_filtered %>%
  group_by(year_new, percent_names, percent_values) %>%
  summarize() %>%
  ungroup() %>%
  arrange(desc(year_new))

plot_waitiing_times_targets_met <- plot_waitiing_times_years_percent_treated(
  waitiing_times_years_percent_treated
)
do_cis_plot(plot_waitiing_times_targets_met, cancer_waiting_times_targets_met_plot_filepath)
```

The dashed red line indicates the 95% target.

* 31 Day targets are being met.
* 62 Day targets are being missed.

The 31 day targets have been met for the last 4 years in a row so this suggests 
there is a 'good enough' system in place, though it could always be better. The 
31 day target relates to reaching the *decision* to treat quickly so this 
likely requires fewer people and resources to review the patient details than 
having to commence delivering the actual treatment.

However, although it was met, the 31 day target was achieved in 2022 but barely 
scraped by on 95.04%.

**The 31 day target is also at high risk of being missed in 2023** and being 
November 2023 at time of writing, little time to put new measures in place.

In the course of this analysis we've become aware that there are places where 
metrics about lab processing of tests such as number of tests successful, 
processing time taken etc. are being recorded. It is encouraging to read in 
screening reports that at various points tests are routinely upgraded to those 
more sensitive or delivering more accurate or quicker results. So it appears 
all measures withing the control of laborotaries are already being done. Do 
clinicians prefer to wait longer to get a more comprehensive set of tests 
completed before deciding on a suitable treatment plan for better longer-term 
outcomes? Does there need to be more laboratory resources for diagnostic tests 
allocated? Are there enough scanners/radiographers? Is the bottleneck more due 
to lower than ideal staffing levels? The latter would likely require 
considerably more financial and time investment.

There seems to be a chronic problem achieving the 62 day target as it has 
not been achieved even as far back as 2012 though other reports indicate it 
*was* achieved in 2011. The target must have been considered achievable when 
it was drawn up. What has changed?

The gap is widening as the years go by except for a brief improvement in 2020 
the year COVID-10 started. If nothing tried so far has worked then this maybe 
needs a more major rethink about how services are allocated and delivered. 


**Suggestion**

Set up a focus group to review what's been tried already and reasons for 
failure to make any real inroads for improvement. It is noted that the NHS in 
England too is experiencing similar missed targets. There they are making 
adjustments to the reporting process"...following a consultation on the cancer 
waiting times standards..." [
Changes to Cancer Waiting Times standards and Official Statistics](https://www.england.nhs.uk/statistics/statistical-work-areas/cancer-waiting-times/). They appear to be about to break categories down into greater 
detail in the data.