---
title: 'Analysis: Death from cancer in Scotland'
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	error = FALSE,
	message = FALSE
)
```

# Death

```{r}
library(dplyr)
library(here)
library(readr)
library(skimr)
library(stringr)
library(tidyr)
# library(tidyverse)
```

```{r}
source(here::here("R/nrs_data_info.R"))
source(here::here("R/death.R"))
# Retrieve helper functions for manipulating cause of death data
source(here::here("R/theming.R"))
```

```{r}
# Set custom theme
theme_set(theme_cancer_in_scotland())
```

```{r}
# Read in clean cancer death data
death_cause <- read_csv(here::here(
  "data_clean", nrs_death_cause_filepath
))
```

```{r}
# View(death_cause)
# Extract list of names of causes of death
causes_of_death <- death_cause %>%
  distinct(cause) %>%
  arrange(str_to_upper(cause)) %>%
  pull()
# causes_of_death

causes_of_death_simple <- death_cause %>%
  arrange(str_to_upper(cause_simple)) %>%
  distinct(cause_simple) %>%
  pull()

# causes_of_death
# Helper variables
label_cause_cancer <- "Cancer"
label_cause_all_causes <- "All causes"
label_cause_non_cancer <- "Non-cancer"

latest_year_nrs_data <- 2022
```
## Background

We wish to see where cancer fits into the overall pattern of death in Scotland.
The [National Records of Scotland](https://www.nrscotland.gov.uk/) (NRS) 
maintain official death statistics for Scotland. 
This analysis utilises NRS data.

**Source:**  
[Table 1: All ages age-standardised death rates for all causes and certain selected causes, Scotland, 1994 to 2022](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/age-standardised-death-rates-calculated-using-the-esp).  
NRS Last update: 19 September 2023.

Definition: `rate` is deaths per 100,000 people in Scotland using. The NRS data is 
 age-standardised using the European Standard Population.

There are `r length(causes_of_death)` categories of causes recorded in the data 
set.

```{r results='asis'}
paste(causes_of_death)
```

As you can see these are formal technical names. For simplifying ongoing work 
the following substitutions have been made and called `cause_simple`.

```{r results='asis'}
paste(causes_of_death_simple)
```

## Analysis

```{r include=FALSE}
dim(death_cause)
# 45 rows 3 columns

# Examine data
glimpse(death_cause)
skimr::skim(death_cause)
```
### How does death from cancer in Scotland compare to other causes?
```{r}
# View(death_cause)
# death_cause

# Create table of available death causes ordered by year high to low to show
# most recent first, then by rate high to low, then alphabetical by cause_simple
death_cause_recent <- death_cause %>%
  select(year, rate, cause_simple) %>%
  mutate(rate = round(rate, 2)) %>%
  arrange(desc(year), desc(rate), cause_simple)
# death_cause_recent
```

```{r}
# #### Cancer death rate as a percentage of death rate from all causes.
death_rate_average <- death_cause_recent %>%
  group_by(cause_simple) %>%
  summarise(average_rate = mean(rate)) %>%
  arrange(desc(average_rate)) %>%
  ungroup()
# death_rate_average

cancer_death_rate_latest <- get_death_by_year_cause_simple(death_cause_recent,
  filter_cause_simple = label_cause_all_causes,
  filter_year = latest_year_nrs_data
)

# cancer_death_rate_latest
cancer_death_rate_cancer_latest <- get_death_by_year_cause_simple(death_cause_recent,
  filter_cause_simple = label_cause_cancer,
  filter_year = latest_year_nrs_data
)
# cancer_death_rate_cancer_latest

cancer_death_rate_cancer_percent <- cancer_death_rate_cancer_latest / cancer_death_rate_latest * 100

cancer_death_proportion <- tibble(
  category = c(label_cause_cancer, label_cause_non_cancer),
  percent = c(cancer_death_rate_cancer_percent, 100 - cancer_death_rate_cancer_percent),
  round(100 - cancer_death_rate_cancer_percent, 1)
) %>%
  mutate(
    percent_labels =
      scales::percent(percent, accuracy = 0.1, scale = 1)
  )

plot_death_rate_cancer(cancer_death_proportion)
```

```{r}
### Selected causes
death_selected_cause_simple <- get_death_selected_cause_simple(death_cause_recent)
#death_cause_simple)
#head(death_selected_cause_simple)

plot_death_selected_cause_simple <- plot_death_cause_simple(death_selected_cause_simple)
plot_death_selected_cause_simple
```

This shows how the **proportion** of deaths is changing over time by cause.  
Cancer is the 3rd bar from the top outlined in black. After 2020, cancer and many other 
non-COVID-19 conditions are once again worsening.

N.B. This **does not** show changes in *amount*. What it does show is the 
relationship between all possible causes. If a particular cause bar is 
getting larger like dementia it is becoming worse or if it is decreasing like circulatory then people are becoming less likely to die from that cause.

For the latter it would indicate change in that a condition that people used to 
die of more in the past they're now dying of *something else* instead, so in 
those cases it's the 'something else' that has worsened and would be the things 
to take especial note of and investigate further. And another interpretation 
can be that treatments other factors are improving things. E.g. we might 
comment that things lke circulatory/heart/cerebrovascular disease have improved 
over the long term considerably since the start of the data in 1994 - could 
this be an example of improved emphasis on preventative work treating thingss
like e.g. High Blood Pressure with medication and lifestyle changes? Is so, 
that would be an encouraging example that change is possible.

A good example is the emergence in 2020 of COVID-19 - a new cause of death marked 
with a dotted red vertical line. People were dying from this new cause instead of 
the kind of thing they would normally have died of. 



#### Table: Rate of death compared to other causes by year
```{r}
# ?DT::datatable
DT::datatable(death_cause_recent,
  caption = "Data for 1994 - 2022"
)
```




```{r}
## All Scotland

### All causes
#death_all_cause_simple <- get_death_all_cause_simple(death_cause_recent)
#death_cause_simple)
#head(death_all_cause_simple)

#plot_death_all_cause_simple <- plot_death_cause_simple(death_all_cause_simple)
#plot_death_all_cause_simple
```
