---
title: "Analysis: Waiting Times for Cancer Treatment"
subtitle: "Cancer in Scotland"
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 5
editor_options: 
  chunk_output_type: inline
---
```{r setup_incidence, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  # error = FALSE,
  message = FALSE
)
```
# Waiting Times for Cancer Treatment

For progressive diseases like cancer early treatment is important to reduce the 
burden of death and disease.

In the publication [Better Cancer Care, An Action Plan](https://www.gov.scot/publications/better-cancer-care-action-plan/), The 
Scottish Government has set targets that once a cancer has been diagnosed 
the patients should start treament within a partcular timeframe.

"...Current standards for cancer waiting times are that 95% of all eligible
patients should wait no longer than 31 or 62 days..."

For this we will be using Public Health Scotland dataset for Cancer Waiting Times.
"Cancer Waiting Times statistics for the 62 day standard for patients urgently 
referred with a suspicion of cancer to first cancer treatment and for the 31 
day standard for patients regardless of the route of referral from date decision 
to treat to first cancer treatment." - Public Health Scotland

Source:  [Cancer Waiting Times](https://www.opendata.nhs.scot/dataset/cancer-waiting-times) [Public Health Scotland]

```{r}
library(janitor)
library(skimr)
library(tidyverse)
```

```{r}
waiting_times_31_days_raw <- read_csv("../data_raw/phs/wait_times/cwt_31_day_standard.csv")
waiting_times_62_days_raw <- read_csv("../data_raw/phs/wait_times/cwt_62_day_standard.csv")
```

```{r eval=FALSE, include=FALSE}
# 31 day waiting standard split by health board of treatment and health board of residence as well as cancer type.
# 
# The 31 day standard is measured against the health board of first treatment (HBT)


#Examine 31 day datsset
dim(waiting_times_31_days_raw)
# 19287    10

# Examine data
glimpse(waiting_times_31_days_raw)
skimr::skim(waiting_times_31_days_raw)

# Lots of NAs but only in description QF fields not values

# Dates are in format "2012Q1", "2012Q1"
# HB = Health Board by code : "S08000015" etc.
# HBT = Health Board of treatment
# CancerType: "Breast", "Cervical"
# NumberOfEligibleReferrals31DayStandard = <dbl> 78, 4, 53
# NumberOfEligibleReferralsTreatedWithin31Days = <dbl> 78, 4, 53

waiting_times_31_days_raw <- waiting_times_31_days_raw %>% 
  janitor::clean_names()
names(waiting_times_31_days_raw)

# [1] "quarter"                                             
#  [2] "hb"                                                  
#  [3] "hbt"                                                 
#  [4] "hbtqf"                                               
#  [5] "cancer_type"                                         
#  [6] "cancer_type_qf"                                      
#  [7] "number_of_eligible_referrals31day_standard"          
#  [8] "number_of_eligible_referrals31day_standard_qf"       
#  [9] "number_of_eligible_referrals_treated_within31days"   
# [10] "number_of_eligible_referrals_treated_within31days_qf"
```
```{r eval=FALSE, include=FALSE}
#Examine 62 day datsset
# 62 day waiting standard split by health board of treatment and health 
# board of residence as well as cancer type
dim(waiting_times_62_days_raw)
# 17362    10

# Examine data
glimpse(waiting_times_62_days_raw)
skimr::skim(waiting_times_62_days_raw)

# Lots of NAs but only in description QF fields not values

# Dates are in format "2012Q1", "2012Q1"
# HB = Health Board by code : "S08000015" etc.
# The 62 day standard is measured against the health board of receipt of referral (HB)

# HBT = Health Board of treatment
# CancerType: "Breast", "Cervical"
# NumberOfEligibleReferrals62DayStandard = <dbl> 52, 2, 31 
# NumberOfEligibleReferralsTreatedWithin62Days = dbl> 52, 2, 29

waiting_times_62_days_raw <- waiting_times_62_days_raw %>% 
  janitor::clean_names()
names(waiting_times_62_days_raw)

#  [1] "quarter"                                             
#  [2] "hb"                                                  
#  [3] "hbqf"                                                
#  [4] "hbt"                                                 
#  [5] "cancer_type"                                         
#  [6] "cancer_type_qf"                                      
#  [7] "number_of_eligible_referrals62day_standard"          
#  [8] "number_of_eligible_referrals62day_standard_qf"       
#  [9] "number_of_eligible_referrals_treated_within62days"   
# [10] "number_of_eligible_referrals_treated_within62days_qf"
```

```{r}
# Lets join these two datasets and rename some columns
# We want to keep everything in the 31day data and any matches on 62 days by
# quarter
# This seems like a 'left' join



```

