---
title: "Analysis: Incidence of cancer in Scotland"
subtitle: "Cancer in Scotland"
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 5
editor_options: 
  chunk_output_type: inline
---
```{r setup_incidence, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  # error = FALSE,
  message = FALSE
)
```

```{r}
library(dplyr)
library(gt)
library(here)
library(readr)
library(stringr)
```

```{r}
source(here::here("R/cancer_in_scotland.R"))
source(here::here("R/incidence.R"))
source(here::here("R/phs_data_info.R"))
source(here::here("R/theming.R"))
```

```{r}
# General settings for this document ----
# Set custom theme
theme_set(theme_cancer_in_scotland()) # base_size=14))
# If TRUE then we will save plots the images folder for use in presentations etc.
```


# Incidence
```{r}
# Read in clean cancer death data
incidence_all <- read_csv(here::here(
  "data_clean", phs_incidence_filepath
))
```

```{r eval=FALSE, include=FALSE}
# skimr::skim(incidence_all)

# N.B this dataset now has over a million rows as it was originally wide and
# contains both numbers and rates

# incidence_all %>%
#  count(cancer_site, sex, year)

# head(incidence_all)
# names(incidence_all)

# [1] "cancer_site"              "cancer_site_icd10code"    "sex"
# [4] "year"                     "incidences_age_range"     "incidences_age"
# [7] "incidence_rate_age_range" "incidence_rate_age"
```

```{r, error=TRUE}
# Important that error=TRUE on this code chunk to stop if invalid ICD-10.

# Before doing the analysis, ensure our dataset has the correct ICD-10 codes
# Do not progress further and alert user that dataset may have changed.
# TODO: remember to uncomment before production
# valid <- check_valid_icd10(incidence_all)
incidence <- simplify_cancer_site(incidence_all)
```

```{r}
incidences_by_year <- get_incidences_by_year(incidence)
lowest_incidences_year <- get_lowest_incidences_by_year(incidences_by_year)
highest_incidences_year <- get_highest_incidences_by_year(incidences_by_year)
lowest_incidences <- min(incidences_by_year$incidences_age)
highest_incidences <- max(incidences_by_year$incidences_age)
oldest_incidences_year <- min(incidences_by_year$year)
newest_incidences_year <- max(incidences_by_year$year)
```

## Background

There are *multiple* different types of cancer. We would like to find out which 
types have been recorded in Scotland and how these have changed over time.

Different cancers are categorised (coded) according to World Health Organisation 
(WHO) [International Statistical Classification of Diseases and Related Health Problems (ICD)](https://www.who.int/standards/classifications/classification-of-diseases).
In Scotland, currently the 'ICD-10' code is used when a patient is diagnosed to 
record which type of cancer they have. Example of an ICD-10 code: 
C50 = ‘Breast’ cancer.

### Terminology

Technical terms used relating to this dataset.

| Term    | Description |
| -------- | ------- |
| **incidence** | The total number of new cases (registrations) of the cancer diagnosed in Scotland for the given period. |
| **cancer site** | The place where a cancer starts in the body. This is represented as a code and also a human readable name. |
----------------------

The [Scottish Cancer Registry and Intelligence Service](https://publichealthscotland.scot/our-areas-of-work/conditions-and-diseases/cancer/scottish-cancer-registry-and-intelligence-service-scris/overview/) (SCRIS) 
collects detailed information on every cancer in Scotland. However, this raw 
information is **not public** and can only be directly accessed by people in the 
NHS and approved individuals.

Public Health Scotland (PHS) makes available extracts from the SCRIS as 
open data.
For this analysis the 
[Annual Cancer Incidence](https://www.opendata.nhs.scot/dataset/annual-cancer-incidence) 
data will be used. Specifically 
[Incidence at Scotland Level](https://www.opendata.nhs.scot/dataset/annual-cancer-incidence/resource/72c852b8-ee28-4fd8-84a9-5f415f4bc325). 
Annual number and rate of new cancer registrations in Scotland by sex and age.

Cancer incidence in Scotland
To December 2021)

The following dataset gives us an annual number and rate of new cancer 
registrations in Scotland by sex and age. Reports include data from 1997 to 2021.

**Source:**  [Incidence at Scotland Level](https://www.opendata.nhs.scot/dataset/c2c59eb1-3aff-48d2-9e9c-60ca8605431d/resource/72c852b8-ee28-4fd8-84a9-5f415f4bc325/download/opendata_inc9721_scotland.csv). 
PHS Last update: 29 March 2023.
 
This analysis utilises PHS data. 
Their sex data uses 'Females' plural but "Male" singular, this has been left to 
match the original.

### Important! When is a medical cancer a 'cancer' in official reports?

"'All cancers' by convention **excludes non-melanoma skin cancers**." - PHS.

Non-melanoma skin cancers are in the original open data supplied. To keep 
consistent with the official reporting standard 'Non-melanoma skin cancers' have
been excluded from this analysis.
-  
[Cancer incidence in Scotland
To December 2021](https://publichealthscotland.scot/publications/cancer-incidence-in-scotland/cancer-incidence-in-scotland-to-december-2021/)

## Analysis

This dataset covers years `r oldest_incidences_year`-`r newest_incidences_year`.

### Time: Cancer Incidences by Year and Sex
```{r}
# Plot INCIDENCE 1, Yearly incidences - a line plot
plot_incidences <- plot_incidences_by_year(incidences_by_year)
do_cis_plot(plot_incidences, cancer_incidence_plot_filepath)
```

The higher the incidences, the more people have been formally diagnosed as 
having a new case of cancer. Shows incidences in general, started increasing in 
number in 1998 to 2018. Then dropped from 2019 to 2020 (start of COVID). In most
recent year of results available 2021 numbers rose sharply to the highest level 
ever.

### Summary: Incidences by Sex

```{r}
# Show the total and mean incidences across years
summary_incidences_by_sex <- incidences_by_year %>%
  group_by(sex) %>%
  summarise(
    total = sum(incidences_age),
    mean = mean(incidences_age),
    median = median(incidences_age)
  ) %>%
  ungroup()

table_summary_incidences_sex(summary_incidences_by_sex,
  lowest_year = oldest_incidences_year,
  highest_year = newest_incidences_year
)
```

On average in Scotland there are almost 30,000 new incidences of cancer each 
year. Females have a higher average(mean) and median number of incidences per 
year than males but bear in mind there is a slightly higher number of females in
the population of Scotland. These figures are relatively close in spite of sexes
having some completely different types of cancer risk due to different organs.

> The 2021 estimates for Scotland's population produced by National Records of 
Scotland (NRS) show that Scotland's overall population stands at 5.48 million: 
2.67 million males and 2.8 million females.  
[Population: estimates, Scotland, Scottish Public Health Observatory](https://www.scotpho.org.uk/population-dynamics/population-estimates-and-projections/data/population-estimates/#:~:text=Scotland,males%20and%202.8%20million%20females.) 

```{r}
incidences_by_year_sex <- incidences_by_year %>%
  # Remove All sexes
  filter(sex != "All")

# Plot INCIDENCE 2, Total incidences by sex - a box plot
plot_total_incidences_sex_boxplot <- # plot_total_incidences_by_year_sex_boxplot(total_incidences_by_sex)
  plot_incidences_boxplot <- plot_incidences_by_year_boxplot(incidences_by_year_sex)
do_cis_plot(plot_total_incidences_sex_boxplot, cancer_incidence_plot_box_plot_filepath)
```
<!-- The lower and upper hinges correspond to the first and third quartiles 
(the 25th and 75th percentiles).  -->

Displays the summary statistics for yearly incidences for both sexes. 
Even taking COVID-19 in to account there are no outliers - no years stand out as 
being significantly different from the others. The median for females is 
higher and males and the distribution is more negatively skewed.

```{r eval=FALSE, warning=FALSE, include=FALSE}
# Model 1 - ignore, just experimenting
# "outcome ~ predictor"
model_year_incidences <- build_model_year_incidences(incidences_by_year_sex)
model_year_incidences
rm(model_year_incidences)
```
```{r}
rm(incidences_by_year_sex)
```

```{r}
# Create a table out of our other variables for rendering as a
# table by gt
incidences_min_max <- tibble(
  result = c(incidence_highest_label, incidence_lowest_label),
  incidences = c(highest_incidences, lowest_incidences),
  year = c(highest_incidences_year, lowest_incidences_year)
)

incidences_min_max <- table_incidences_min_max(
  incidences_min_max,
  oldest_incidences_year, newest_incidences_year
)
incidences_min_max
```

Cancer incidences had been steadily increasing since the late 1990s before starting 
to drop in 2019, note this was the year *before* COVID-19 started. Unfortunately the
most recent year of data the cases are at their highest number ever. This has to 
be caveated with our growing and ageing population. So the drop on the
year of COVID-19 itself might have been part of an already decreasing trend. There
was widespread disruption to health services and daily life. All manner 
of primary care (e.g. GP practices) and diagnostic services were interrupted 
completely or on heavily reduced capacity which may give some administrative 
reasons for a drop in formal diagnoses. The steep rise 2020-2021 may have a 
backlog of previously undiagnosed cases as part of that increase.

GPs commented at the time that a lot of patients weren't coming to them. they are often the first point of contact for those with symptoms/concerns eventually leading to a formal cancer diagnosis.
["GPs are worried about “missing patients” and the backlog which has built up in chronic disease management, screening, immunisation and cancer referrals." DEEP END REPORT 36 General Practice in the time of Covid-19](https://www.gla.ac.uk/media/Media_728030_smxx.pdf) June 2020.

People with weak immune systems were more vulnerable to negative outcomes from 
the virus and the elderly being more likely to die from COVID-19 may also have 
had the side effect of reducing the number of old people with a diagnosis of a 
new case of cancer. This is probably the sort of question that might be able to 
be tested by those with access to the Scottish Cancer Registry to track 
patients with other multiple conditions.

### Types of Cancer (Cancer Sites)
```{r}
incidences_by_cancer_site <- get_incidences_by_cancer_site(incidence)
```

#### Abbreviations for charts

Some of the names of the cancers are rather long. To help make these 
more legible on charts the following substitutions have been made. If 
you see "..." next to a name it means it's been adjusted to use the names 
in the following table.

```{r}
incidences_shortened_cancer_site <- get_shortened_cancer_site(incidences_by_cancer_site)
# Show a table of abbreviations
table_shortened_cancer_site(incidences_shortened_cancer_site)
```

```{r}
incidences_cancer_site <- get_cancer_site(incidences_by_cancer_site)
incidences_number_of_cancer_sites <- nrow(incidences_cancer_site)
```
#### Incidences by Year and Cancer Sites

In Scotland cancers are categorised (coded) to World Health Organisation 
standards.

Here are the **recorded*** cancers for the all the years available (`r oldest_incidences_year`-`r newest_incidences_year`) in the dataset. 

There are **`r incidences_number_of_cancer_sites`** types of cancer (cancer 
sites). Some of these names are themselves umbrella categories of other more specific sub-categories e.g. 'Leukaemias' and 'Colorectal cancer' (Colon+Rectum) might be better known by the public as 'Bowel' cancer.

These represent multiple years and all ages and sexes.  
* Some excluded from all cancer reporting e.g. non-malignant skin cancer.

```{r  fig.height = 8, fig.width = 10}
# Plot INCIDENCE 3, incidences_by_year_cancer_site
plot_incidences_bycs <- plot_incidences_by_year_cancer_site(
  incidences_by_cancer_site, oldest_incidences_year, newest_incidences_year
)
do_cis_plot(plot_incidences_bycs, cancer_incidence_by_year_cancer_sites_plot_filepath)
```

##### All time (`r oldest_incidences_year`-`r newest_incidences_year`)

```{r}
# Table showing highest and lowest incidences and which years
table_incidences_cancer_sites_between(
  incidences_cancer_site,
  oldest_incidences_year, newest_incidences_year
)
```
```{r}
# Lung, Breast, Bowel and Prostate
largest_types <- ((125189 + 107964 + 93160 + 75673) / 749411) * 100
# =  53.6402588
rm(largest_types)
```

Lung, Breast, Bowel and Prostate alone account for over 53% of the total.

##### Most recent year available (`r highest_incidences_year`)

For comparison here is the list only for the latest year of data.

```{r}
incidences_cancer_site_recent <- incidences_by_cancer_site %>%
  select(cancer_site, incidences_age, year) %>%
  group_by(cancer_site) %>%
  filter(year == highest_incidences_year) %>%
  summarise(total_incidences = sum(incidences_age)) %>%
  arrange(desc(total_incidences)) %>%
  ungroup()
```

```{r}
# Table showing highest and lowest incidences and which years
table_incidences_cancer_sites(incidences_cancer_site_recent, oldest_incidences_year, newest_incidences_year)
```

```{r}
# Trying to find some way of comparing the two incidence dataframes
# Im imagining something like a chart with up or down arrows to show
# movement
# min_rank(desc(cancer_site_recent$total_incidences))
```

What might be worth watching is which cancer sites are in a higher position in 
the 2021 table as this may show cancers that are starting to become more 
prevalent. e.g. "Malignant melanoma of the skin" and "Head and neck" have moved 
to positions ahead of "Rectum and rectosigmoid junction" and "Kidney" now makes 
the top 10 whereas "Non-Hodgkin lymphoma" drops to 11th.

### Age Range

What age are people when they are newly diagnosed. Would knowing the 
demographics of different diseases affect what kind of public health information
and awareness raising needs to be done? The next tables are based on all of the 
years of data available.

```{r}
# Get Cancers incidences by age range and sex
incidences_all_cancer_site_age_range <- get_cancer_site_age_range(incidence)
```

#### Top Incidences by Cancer Site, Age Range and Sex

This is a list of those cancer sites, age groups and sexes most at risk of 
having a new cancer diagnosed. 

```{r}
# Show top 10 across all
table_age_range <- get_top_incidences_age_range_sex(
  incidences_all_cancer_site_age_range,
  top = 10
)
table_incidences_cancer_sites_age_range(table_age_range, "Top 10 All")
```

Next an example of how different the top risks by cancer site can be by age 
and sex.

```{r}
# Top 5 by Females 20-29
table_age_range <- get_top_incidences_age_range_sex(incidences_all_cancer_site_age_range,
  top = 5,
  filter_sex = "Females",
  filter_age =
    c("20-24", "25-29")
)
table_incidences_cancer_sites_age_range(table_age_range, "Top 5 Females 20-29")
```

Next, an example of what age range and sex might be more at risk of a 
specific cancer type.

```{r}
# Top 5 by cancer site "Hodgkin lymphoma"
table_age_range <- get_top_incidences_age_range_sex(incidences_all_cancer_site_age_range,
  top = 5,
  filter_sex = NULL,
  filter_cancer_site = "Hodgkin lymphoma"
)
table_incidences_cancer_sites_age_range(table_age_range, "Top 5 Hodgkin lymphoma")
rm(table_age_range)
``` 

#### Incidences by Age Range, Sex and Cancer Site

Here we have age-related data where incidences have been provided in 5-year 
categories  from 0, .. to 90+. This data gives us the chance to consider the 
relationship between ageing and risk of being diagnosed with a new cancer.

```{r  fig.height = 8, fig.width = 10}
# Plot INCIDENCE 4, incidences_all_cancer_site_age_range
plot_age_range <- plot_all_cancer_site_age_sex_range(incidences_all_cancer_site_age_range)
do_cis_plot(plot_age_range, cancer_incidence_by_age_range_plot_filepath)
```

We see 4 main cancers dominating in terms of numbers Colorectal/Lung/Breast and 
Prostate. The pattern over many cancers in terms of age range is that most of 
them really become much more prevalent from middle age onwards, peaking for new 
cases for most sites when we are in our 70s. 

Of these big 4 "Breast" seems 
highest at younger ages than the other 3 so awareness raising should start in 
younger age groups. The ones that buck that trend higher more in the first half 
of life tend to be the ones associated with "Cervix uteri" (Cervical) in females
and "Testis" (Testicular) in males.

**Suggestion:**

* Is the public aware of the increased risk simply by ageing? Are there 
opportunities to raise awareness among people engaging with other health or 
public sector organisations to be watchful for early symptoms that might be 
higher risks according to the age of the person.

* There are already screening programmes for breast/bowel and cervical cancer.
Lung and Prostate cancer are some of the top in terms of numbers. Is there some equivalent of screening for these two in particular? If not, keep informed of any developments in early detection of these two.

## Project Output: Cancer Risk Tool

In the course of this analysis an interactive Cancer Risk Tool has been produced that allows users to experiment with changing age/sex and cancer site 
and getting an idea of what the risk of that combination might be. This is 
leveraging the value of a multi-decade dataset in capturing risk in Scotland.

**Technical Note:**

The tool will be found in the "cancer_risk_tool" folder that accompanies this report. It is an experimental demo and requires the user to run the *Shiny R Markdown* document by opening the file and clicking "Run Document".

Here is a screenshot of the tool in action

TODO screenshot

***
```{r}
# rm(list = ls(pattern = "incidences_"))
# rm(list = ls(pattern = "incidence"))
```
