---
title: "Analysis: incidence by type of cancer in Scotland"
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 4
---
```{r setup_incidence, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  error = FALSE,
  message = FALSE
)
```

```{r}
source(here::here("R/phs_data_info.R"))
source(here::here("R/theming.R"))
```

```{r}
# Set custom theme
theme_set(theme_cancer_in_scotland())
```



# Incidence
## Background

Cancer is a category of multiple different types of cancer. We would like to 
find out which types have been recorded in Scotland and how these have changed 
over time
We want where cancer fits into the overall pattern of death in Scotland.

The Scottish Cancer Registry (SMR06) collates this information but it can only 
be directly accessed by people in the NHS.

Terminology: `cancer site` - the place where a cancer starts in the body

Public Health Scotland (PHS) makes available aggregated opendata extracted from 
the registry via 
[Annual Cancer Incidence](https://www.opendata.nhs.scot/dataset/annual-cancer-incidence)

The following dataset gives us an nnual number and rate of new cancer 
registrations in Scotland by sex and age.

**Source:**  [Incidence at Scotland Level](https://www.opendata.nhs.scot/dataset/c2c59eb1-3aff-48d2-9e9c-60ca8605431d/resource/72c852b8-ee28-4fd8-84a9-5f415f4bc325/download/opendata_inc9721_scotland.csv)
PHS Last update: 29 March 2023
 
This analysis utilises PHS data.

## Analysis

```{r}
#incidence_raw <- read_csv()
# Read in clean cancer death data
incidence <- read_csv(here::here(
  "data_clean", phs_incidence_filepath
))
```

