---
title: 'Analysis: Death from cancer in Scotland'
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 4
#    toc_float: true
   
  df_print: paged
---
```{r setup_death, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  error = FALSE,
  message = FALSE
)
```

# Death

```{r}
library(dplyr)
library(here)
library(readr)
library(skimr)
library(stringr)
library(tidyr)
# library(tidyverse)
```

```{r}
source(here::here("R/nrs_data_info.R"))
source(here::here("R/death.R"))
# Retrieve helper functions for manipulating cause of death data
source(here::here("R/theming.R"))
```

```{r}
# Set custom theme
theme_set(theme_cancer_in_scotland())
```

```{r}
enable_plot_saving
```


```{r}
# Read in clean cancer death data
death_cause <- read_csv(here::here(
  "data_clean", nrs_death_cause_filepath
))
```

```{r}
# View(death_cause)
# Extract list of names of causes of death
causes_of_death <- death_cause %>%
  arrange(str_to_upper(cause)) %>%
  distinct(cause)

causes_of_death_simple <- death_cause %>%
  arrange(str_to_upper(cause_simple)) %>%
  distinct(cause_simple)

# causes_of_death
# Helper variables
label_cause_cancer <- "Cancer"
label_cause_all_causes <- "All causes"
label_cause_non_cancer <- "Non-cancer"

latest_year_nrs_data <- 2022
```
## Background

We wish to see where cancer fits into the overall pattern of death in Scotland.
The [National Records of Scotland](https://www.nrscotland.gov.uk/) (NRS) 
maintain official death statistics for Scotland. 
This analysis utilises NRS data.

When a death is registered, it is common for a number of diseases or conditions 
tobe recorded on the death certificate. NRS apply the WHO rules to determine 
which of these conditions is the underlying cause of death." - 
[Death Certificates and Coding the Causes of Death](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/deaths-background-information/death-certificates-and-coding-the-causes-of-death)

**Source:**  
[Table 1: All ages age-standardised death rates for all causes and certain selected causes, Scotland, 1994 to 2022](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/age-standardised-death-rates-calculated-using-the-esp).  
NRS Last update: 19 September 2023.

Definition: `rate` is deaths per 100,000 people in Scotland using. The NRS data 
is age-standardised using the European Standard Population.

There are `r count(causes_of_death) %>% pull()` categories of causes recorded
in the dataset.

```{r}
causes_of_death %>% pull()
```

As you can see these are formal technical names. For simplifying ongoing work 
the following substitutions have been made and called `cause_simple`.

```{r}
causes_of_death_simple %>% pull()
```

## Analysis

```{r include=FALSE}
dim(death_cause)
# 45 rows 3 columns

# Examine data
glimpse(death_cause)
skimr::skim(death_cause)
```
### How does death from cancer in Scotland compare to other causes?
#### Overview
```{r}
# View(death_cause)
# death_cause

# Create table of available death causes ordered by year high to low to show
# most recent first, then by rate high to low, then alphabetical by cause_simple
death_cause_recent <- death_cause %>%
  select(year, rate, cause_simple) %>%
  mutate(rate = round(rate, 2)) %>%
  arrange(desc(year), desc(rate), cause_simple)
# death_cause_recent
```

```{r}
# #### Cancer death rate as a percentage of death rate from all causes.
death_rate_average <- death_cause_recent %>%
  group_by(cause_simple) %>%
  summarise(average_rate = mean(rate)) %>%
  arrange(desc(average_rate)) %>%
  ungroup()
# death_rate_average

cancer_death_rate_latest <- get_death_by_year_cause_simple(death_cause_recent,
  filter_cause_simple = label_cause_all_causes,
  filter_year = latest_year_nrs_data
)

# cancer_death_rate_latest
cancer_death_rate_cancer_latest <- get_death_by_year_cause_simple(death_cause_recent,
  filter_cause_simple = label_cause_cancer,
  filter_year = latest_year_nrs_data
)
# cancer_death_rate_cancer_latest

cancer_death_rate_cancer_percent <- cancer_death_rate_cancer_latest / cancer_death_rate_latest * 100

cancer_death_proportion <- tibble(
  category = c(label_cause_cancer, label_cause_non_cancer),
  percent = c(cancer_death_rate_cancer_percent, 100 - cancer_death_rate_cancer_percent),
  round(100 - cancer_death_rate_cancer_percent, 1)
) %>%
  mutate(
    percent_labels =
      scales::percent(percent, accuracy = 0.1, scale = 1)
  )

plot_death_rate <- plot_death_rate_cancer(cancer_death_proportion,
  plot_title = paste(
    plot_death_cause_simple_title,
    "\nLatest year:", latest_year_nrs_data
  )
)
plot_death_rate
if (enable_plot_saving) {
  ggsave(
    here::here(plot_death_cause_simple_plot_filepath),
    plot_death_rate
  )
}
```

This shows the proportion of deaths from cancer compared to all other 
causes combined for the most recently available year of data - `r latest_year_nrs_data`.

At time of writing cancer accounts for almost exactly a quarter of all deaths.

#### Change over time for all causes of death

```{r}
### Selected causes
death_selected_cause_simple <- get_death_selected_cause_simple(death_cause_recent)
# death_cause_simple)
# head(death_selected_cause_simple)

plot_death_selected_cause_simple <- plot_death_cause_simple(death_selected_cause_simple)
plot_death_selected_cause_simple
```

This shows how the **proportion** of deaths is changing over time by cause.  
Cancer is the thirds bar from the top. A dotted red vertical line 
marks 2020 -the start of the COVID-19 pandemic. After 2020 many other 
non-COVID-19 conditions are once again worsening, 
however, **the rate of cancer deaths is improving**, albeit in a very small way.
You can see this in the precise numbers if you filter the following table by 
'Cancer'.

N.B. This **does not** show changes in *amount*. What it does show is the 
relationship between all possible causes. If a particular cause bar is 
getting larger like dementia it is becoming worse or if it is decreasing like circulatory then people are becoming less likely to die from that cause.

For the latter it would indicate change in that a condition that people used to 
die of more in the past they're now dying of *something else* instead, so in 
those cases it's the 'something else' that has worsened and would be the things 
to take especial note of and investigate further. 

A good example is the emergence in 2020 of COVID-19. People were dying from this
new cause instead of the kind of thing they would normally have died of, 
including cancer. People with cancer and due to certain treatments can have 
weaker immune systems and be more vulnerable to disease. The bar for cancer 
certainly seems to shrink more than other causes in 2020. A further suggestion 
would be to investigate if more people with Cancer and/or Respitory were more 
likely to die if they subsequently caught COVID.

And another interpretation 
can be that treatments other factors are improving things. E.g. we might 
comment that things lke circulatory/heart/cerebrovascular disease have improved 
over the long term considerably since the start of the data in 1994 - could 
this be an example of improved emphasis on preventative work treating thingss
like e.g. High Blood Pressure with medication and lifestyle changes? Is so, 
that would be an encouraging example that change is possible.

#### Table: Rate of death compared to other causes by year

**Tip:** To filter by a particular condition type in the 'Search' box.

```{r}
# ?DT::datatable
DT::datatable(death_cause_recent,
  caption = "Data for years 1994 - 2022"
)
```

#### Recent change in rate

```{r}
# Let's check which causes changed most in the most recent year
recent_change <- death_cause_recent %>%
  # select(cause_simple) %>%
  group_by(cause_simple) %>%
  filter(year > latest_year_nrs_data - 2) %>% # Lest couple of years available
  mutate(rate_previous = lead(rate)) %>%
  filter(year == latest_year_nrs_data) %>%
  mutate(
    rate_change = round(rate - rate_previous, 1),
    rate_change_pct = round(rate_change / rate * 100, 1)
  ) %>%
  ungroup()

# recent_change ?datatable(r
recent_change %>%
  DT::datatable(
    caption = paste("Data for latest year available ", latest_year_nrs_data),
    options = list(searching = FALSE, paging = FALSE)
  )
```
##### Improving since previous year available - `r latest_year_nrs_data-1`

These are what got 
'better' in terms of death rate between the latest two years of data.


```{r}
best_rate_change <- recent_change %>%
  filter(rate_change_pct < 0) %>%
  arrange(rate_change_pct) %>%
  select(cause_simple, rate_change_pct)

DT::datatable(best_rate_change,
  options = list(searching = FALSE, paging = FALSE)
)
```

Conclusion: **rates of cancer deaths have improved for the most recent year**.

It's not possible to tell from our data whether any of the screening or 
vacination programmes contribute in any way to the improvement. And the 
improvement is so small it's almost static.

Of interest is that in spite of a great deal of negative publicity 
about Scottish drug related deaths being worse than many other countries there 
*has* been an improvement recently.

##### Worsening since previous year available - `r latest_year_nrs_data-1`

Ordered by greatest percentage change (worst) to least.

```{r}
worst_rate_change <- recent_change %>%
  filter(rate_change_pct > 0) %>%
  arrange(desc(rate_change_pct)) %>%
  select(cause_simple, rate_change_pct)
# worst_rate_change

DT::datatable(worst_rate_change,
  options = list(searching = FALSE, paging = FALSE)
)
```



This category includes things 
like flu and pneumonia. There is a flu vacination programme in Scotland, do they 
have more data available that could be checked? 

**Conclusion: Respiratory conditions have worsened the most**.

Keep an eye on issues relating to Respiratory and Pulmanory as these worsened the most in the last two years.

**Conclusion: Preventable causes of death like alcohol and suicide have risen**.

More work needs to be done to target these.

```{r}
## All Scotland

### All causes
# death_all_cause_simple <- get_death_all_cause_simple(death_cause_recent)
# death_cause_simple)
# head(death_all_cause_simple)

# plot_death_all_cause_simple <- plot_death_cause_simple(death_all_cause_simple)
# plot_death_all_cause_simple
```
