---
title: 'Cancer Screening programmes in Scotland'
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 4
    df_print: paged
---
```{r setup_screening, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  error = FALSE,
  message = FALSE
)
```
# Screening

```{r}
library(forcats)
library(gt)
library(here)
library(skimr)
library(tidyverse)
```

```{r}
source(here::here("R/phs_data_info.R"))
source(here::here("R/screening.R"))
source(here::here("R/theming.R"))
# Retrieve helper functions for manipulating bowel cancer screening data
```
```{r}
# Set custom theme
theme_set(theme_cancer_in_scotland())
```
```{r}
screening_bowel_cancer_screening_update <- "Bowel Cancer Screening Uptake"
```

## Scottish Bowel Screening Programme

Commenced in 2007 with a full rollout across all health boards by 2009.

People aged 50-74 years are invited every two years to complete a home bowel 
screening test. 75 and over can self-refer. 

### Uptake

The [Healthcare Improvement Scotland
(HIS) standard target](https://www.healthcareimprovementscotland.org/our_work/standards_and_guidelines/stnds/bowel_screening_standards.aspx) for bowel screening is **uptake is 60%**. 

Based on [Scottish bowel screening programme statistics For the period of invitations from May 2020 to April 2022](https://publichealthscotland.scot/publications/scottish-bowel-screening-programme-statistics/scottish-bowel-screening-programme-statistics-for-the-period-of-invitations-from-may-2020-to-april-2022/)

"The proportion of invites which are returned as a correctly
completed kit (i.e. a positive or negative test result). It is a key measure of the
acceptability of the test to the public."

Health Improvement Scotland [Bowel screening standards: August 2023 (PDF)](https://www.healthcareimprovementscotland.org/his/idoc.ashx?docid=49b039aa-f6cf-42aa-8fda-3d6bc165bd21&version=-1).

```{r}
screening_bowel_cancer_takeup_all <- read_csv(here::here(
  "data_clean", phs_screening_bowel_uptake_filepath
))
```

```{r eval=FALSE, include=FALSE}
dim(screening_bowel_cancer_takeup_all)
# 45 rows 3 columns

# Examine data
glimpse(screening_bowel_cancer_takeup_all)
skimr::skim(screening_bowel_cancer_takeup_all)

# The area column represents both health boards and all of Scotland so be
# careful to filter our results if e.g. only health boards requires
```

```{r eval=FALSE, include=FALSE}
# View(screening_bowel_cancer_takeup)
head(screening_bowel_cancer_takeup_all)
```

## All Scotland

```{r}
screening_caption_screening_bowel_cancer_takeup <- 
  "Data for May 2020 to April 2022"
screening_bowel_cancer_takeup_scotland <- get_screening_bowel_cancer_takeup_scotland(screening_bowel_cancer_takeup_all)
```

```{r}
# Draw bar chart
plot_screening_bowel_cancer_takeup <- plot_screening_bowel_cancer_takeup_scotland(screening_bowel_cancer_takeup_scotland)
plot_screening_bowel_cancer_takeup
ggsave(
  here::here(screening_bowel_cancer_uptake_plot_filepath),
  plot_screening_bowel_cancer_takeup
)
```
The 
red dotted horizontal line marks the Health Improvement Scotland standard **target 
of `r screening_bowel_cancer_uptake_target`%**.

This shows that in the current Bowel Cancer Screening latest results around 2/3 
of the people invited via the screening programme are taking up the offer. 

Females are more likely to return the tests than males.
```{r}
# Table of results
get_screening_bowel_cancer_takeup_scotland <- function(df,
           table_title = screening_bowel_cancer_screening_update,
           table_subtitle = screening_caption_screening_bowel_cancer_takeup) {
df %>%
    gt() %>%
    tab_header(
      title = table_title,
      subtitle = table_subtitle
    ) %>%
    cols_label(
      uptake_pct = html("Uptake (%)"),
      sex = html("Sex"),
      area = html("Health Board"),
    ) %>%
    tab_source_note(source_note = source_phs) %>%
    fmt_number(columns = c(uptake_pct), decimals = 1)
}
get_screening_bowel_cancer_takeup_scotland(screening_bowel_cancer_takeup_scotland)
```

For the most recent round of testing the target has been met and exceeded by a
clear margin.

Continue to enhance communication to try and engage the remaining third of the 
general public making more efforts to target males.

**Conclusion: `r screening_bowel_cancer_uptake_target`% target is being met but room for improvement**.

## All Uptake by Health Board and Sex

```{r}
screening_bowel_cancer_uptake_all_means <- screening_bowel_cancer_takeup_all %>%
  group_by(area, sex, uptake_pct) %>%
  summarize(value = mean(uptake_pct))

plot_screening_bowel_cancer_uptake_all <- function(df) {
  plot_screening <- df %>%
    filter(sex != "All persons") %>%
    ggplot(aes(
      x = fct_reorder(area, value),
      y = uptake_pct,
      fill = sex
    )) +
    geom_col(position = "dodge") +
    scale_fill_cis_qualitative(cis_palette_sex) +
    scale_y_continuous(limits = c(0, 100)) +
    coord_flip() +
    labs(
      caption = source_phs,
      x = "Health Board",
      y = "",
      fill = "Sex"
    )
  return(plot_screening)
}

plot_screening_bowel_cancer_uptake_all(screening_bowel_cancer_uptake_all_means)
rm(screening_bowel_cancer_uptake_all_means)
```

Females return their testing kits more than males for every Health Board in Scotland. 

Just to confirm the numbers for each Health Board the following table is 
ordered by those with the lowest uptake first.

```{r}
# Get data frame of health board and sex related data
screening_bowel_cancer_takeup_health_boards <-
  get_screening_bowel_cancer_takeup(
    screening_bowel_cancer_takeup_all,
    # filter_sex = c("Females","Males"),
    filter_sex = c("All persons"),
    # filter_area = c("Grampian", "Fife"),
    remove_scotland = TRUE,
    order_by_pct = TRUE
  )
get_screening_bowel_cancer_takeup_scotland(screening_bowel_cancer_takeup_health_boards)
```
## Conclusion

* Highest: Shetland (`r round(max(screening_bowel_cancer_takeup_health_boards$uptake_pct), digits = 1)`%) 
* Lowest: Greater Glasgow and Clyde (`r round(min(screening_bowel_cancer_takeup_health_boards$uptake_pct), digits = 1)`%) 

There is almost 12% difference between the highest and lowest uptakes between
health boards, this seems seems relatively large. In general the further the 
health board is from the Central Belt of Scotland the better the public 
participation.

Other research suggests that there may be a link in general between deprivation
and uptake of screening programmes. In Scotland the Bowel Cancer screening kits
are sent directly to the home of a person and there is no charge for posting 
them to return the kit. If deprivation is a factor, superficially it wouldn't 
seem like it would be for reasons of cost, so perhaps there are more complex 
reasons around education and awareness of importance or others that need 
teasing out.

**Suggestion:**

There should be some greater examination of what other health 
boards can learn about the roll out in more successful areas like Shetland.
In general the further away from the central belt, the more successful. 
And more efforts to educate and inform the public, particularly males over 60.

Try the interactive Cancer Risk Tool that 
accompanies the analysis report on incidences
with the settings **(All age ranges, Male, Colorectal cancer) **

```{r}
# screening_bowel_cancer_takeup_kpi <-
#  get_screening_bowel_cancer_takeup_kpi1(screening_bowel_cancer_takeup)
# screening_bowel_cancer_takeup_kpi
```

```{r}
rm(list = ls(pattern = "screening"))
```

