---
title: 'Analysis: Death from cancer in Scotland'
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
---
```{r setup, include=FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```
# Death

```{r}
library(dplyr)
library(here)
library(readr)
library(skimr)
library(stringr)
library(tidyr)
# library(tidyverse)
```

```{r}
source(here::here("R/nrs_data_info.R"))
source(here::here("R/death.R"))
# Retrieve helper functions for manipulating cause of death data
```

```{r}
# Read in clean cancer death data
death_cause <- read_csv(here::here(
  "data_clean", nrs_death_cause_filepath
))
```

```{r}
# View(death_cause)
# Extract list of names of causes of death
causes_of_death <- death_cause %>%
  distinct(cause) %>%
  arrange(str_to_upper(cause)) %>%
  pull()
# causes_of_death

causes_of_death_simple <- death_cause %>%
  arrange(str_to_upper(cause_simple)) %>%
  distinct(cause_simple) %>%
  pull()

# causes_of_death
# Helper variables
label_cause_cancer <- "Cancer"
label_cause_all_causes <- "All causes"
label_cause_non_cancer <- "Non-cancer"

latest_year_nrs_data <- 2022

theme_set(theme_minimal())
```
## Background

We wish to see where cancer fits into the overall pattern of death in Scotland.
The [National Records of Scotland](https://www.nrscotland.gov.uk/) (NRS) 
maintain official death statistics for Scotland. 
This analysis utilises NRS data.

**Source:**  
[Table 1: All ages age-standardised death rates for all causes and certain selected causes, Scotland, 1994 to 2022](https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/vital-events/deaths/age-standardised-death-rates-calculated-using-the-esp).  
NRS Last update: 19 September 2023.

Definition: `rate` is deaths per 100,000 people in Scotland using. The NRS data is 
 age-standardised using the European Standard Population.

There are `r length(causes_of_death)` categories of causes recorded in the data 
set.

```{r results='asis'}
paste(causes_of_death)
```

As you can see these are formal technical names. For simplifying ongoing work 
the following substitutions have been made and called `cause_simple`.

```{r results='asis'}
paste(causes_of_death_simple)
```

## Analysis

```{r include=FALSE}
dim(death_cause)
# 45 rows 3 columns

# Examine data
glimpse(death_cause)
skimr::skim(death_cause)
```
### How does death from cancer in Scotland compare to other causes?
```{r}
# View(death_cause)
# death_cause

# Create table of available death causes ordered by year high to low to show
# most recent first, then by rate high to low, then alphabetical by cause_simple
death_cause_recent <- death_cause %>%
  select(year, rate, cause_simple) %>%
  mutate(rate = round(rate, 2)) %>%
  arrange(desc(year), desc(rate), cause_simple)
# death_cause_recent
```

```{r}
# #### Cancer death rate as a percentage of death rate from all causes.
death_rate_average <- death_cause_recent %>%
  group_by(cause_simple) %>%
  summarise(average_rate = mean(rate)) %>%
  arrange(desc(average_rate)) %>%
  ungroup()
# death_rate_average

cancer_death_rate_latest <- get_death_by_year_cause_simple(death_cause_recent,
  filter_cause_simple = label_cause_all_causes,
  filter_year = latest_year_nrs_data
)

# cancer_death_rate_latest
cancer_death_rate_cancer_latest <- get_death_by_year_cause_simple(death_cause_recent,
  filter_cause_simple = label_cause_cancer,
  filter_year = latest_year_nrs_data
)
# cancer_death_rate_cancer_latest

cancer_death_rate_cancer_percent <- cancer_death_rate_cancer_latest / cancer_death_rate_latest * 100

cancer_death_proportion <- tibble(
  category = c(label_cause_cancer, label_cause_non_cancer),
  percent = c(cancer_death_rate_cancer_percent, 100 - cancer_death_rate_cancer_percent),
  round(100 - cancer_death_rate_cancer_percent, 1)
) %>%
  mutate(
    percent_labels =
      scales::percent(percent, accuracy = 0.1, scale = 1)
  )

# cancer_death_proportion %>%
#   # Basic piechart with percentage labels
#   ggplot(aes(x = "", y = percent, fill = category)) +
#   geom_col(color = "white", width = 1) +
#   geom_label(aes(label = percent_labels),
#     position = position_stack(vjust = 0.5),
#     show.legend = FALSE # hide the 'a' symbol
#   ) +
#   #   scale_fill_manual(values = c("red", "yellow")) +
# 
#   coord_polar(theta = "y") + # ?coord_polar
#   labs(
#     title = "Death rate from cancer as percentage of death rate all causes",
#     x = "",
#     y = "Percent",
#     fill = ""
#   )
plot_death_rate_cancer(cancer_death_proportion)
```

#### Table: Rate of death compared to other causes
```{r}
# ?DT::datatable
DT::datatable(death_cause_recent,
  caption = "Data for 1994 - 2022"
)
```



## All Scotland
### All causes

```{r}
death_all_cause <- get_death_all_cause(death_cause)
head(death_all_cause)

plot_death_all_cause <- plot_death_cause(death_all_cause)
plot_death_all_cause
```

### Selected causes

```{r}
death_selected_cause <- get_death_selected_cause(death_cause)
head(death_selected_cause)

plot_death_selectect_cause <- plot_death_cause(death_selected_cause,
  plot_title = "Death rates for selected causes"
)
plot_death_selectect_cause
```
